---
title: "COVID Data Investigation Report"
author: "Chris Toth"
date: "2022-06-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

With this analysis, I intend to investigate how population density affect COVID case rates.  With a disease like COVID which undergoes respiratory transmission, I would expect to see areas with higher population density to exhibit higher case rates and mortality rates.  To conduct this analysis, I will normalize total cases and deaths to population to obtain case rates, and normalize populations to obtain population density. 


## Data Source

The source for the COVID data and a description of what the data is can be found here:
<https://github.com/CSSEGISandData/COVID-19>.

Population data was acquired from the US Census Bureau at:
<https://www.census.gov/data/datasets/time-series/demo/popest/2010s-state-total.html#par_textimage_1873399417>
(download file: nst-est2019-alldata.csv)

State areas were acquired from: 
<https://github.com/jakevdp/data-USstates/blob/master/state-areas.csv>

## Import Data
```{r import, echo=TRUE, results='hide',message=FALSE}
library(tidyverse)
library(RCurl)

covid_data_file <- getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports_us/06-18-2022.csv")
covid_data <- read_csv(covid_data_file)
state_pop_file <- getURL("https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/national/totals/nst-est2019-alldata.csv")
state_pops <- read_csv(state_pop_file)
area_data_file <- getURL("https://raw.githubusercontent.com/jakevdp/data-USstates/master/state-areas.csv")
area_data <- read_csv(area_data_file)
```

The raw, untidy data is initially divided across 3 tables, one for covid data by state, one for state land area, and another for state populations:

```{r raw_data, echo=TRUE, message=TRUE}
covid_data
state_pops
area_data
```

## Tidy and Transform Data

Prepare data for analysis by:
 -  Converting columns to appropriate data types
 -  Dropping unnecessary columns
 -  Joining tables on State name
 -  Sanitize column names
 
Then normalize cases and deaths to population and normalize population to land area, so we have cases and deaths per capita and population density. 
 
No further transformation will be applied at this point because because each row will correspond to an individual observation and each column will correspond to a parameter.


```{r tidy_and_join, echo=TRUE,message=FALSE}
state_pops <- state_pops %>% 
  mutate(STATE = NAME) %>%
  mutate(pop_2019 = POPESTIMATE2019) %>%
  select(c(STATE,pop_2019))

covid_data <- covid_data %>% 
  select(c(Province_State,Confirmed,Deaths)) %>%
  rename(STATE = Province_State)  

area_data <- area_data %>% 
  rename(SQMI = `area (sq. mi)`) %>%
  rename(STATE = state)

covid_data <- covid_data %>%
    inner_join(area_data,by = "STATE") %>%
    inner_join(state_pops,by = "STATE")

covid_data <- covid_data %>%
    mutate(CASES_PC = Confirmed/pop_2019) %>%
    mutate(DEATHS_PC = Deaths/pop_2019) %>%
    mutate(POP_DENS = pop_2019/SQMI) %>%
    select(c(STATE,CASES_PC,DEATHS_PC,POP_DENS))

covid_data <- covid_data %>%
    mutate(LOG_POP_DENS = log10(covid_data$POP_DENS)) %>%
    mutate(MORT_RATE = DEATHS_PC/CASES_PC)

```

The log of population was also computed to to better utilize the dynamic range in subsequent colorbars, but was not found to be useful in any quantitative analysis

The final data, after transformations looks as follows:
```{r tidy_disp, echo=FALSE, message=TRUE}
covid_data
```

## Controls on mortality rate

For a preliminary investigation of how the key variables relate to each other, I made the following scatter plots:

```{r casesVdeaths_graph, echo=TRUE,message=FALSE}
ggplot(data = covid_data) + geom_point(mapping = aes(x = CASES_PC, y = DEATHS_PC, color = LOG_POP_DENS))
```

There is a general positive correlation between deaths rate and case rate, but neither of these variables seems to have a very positive correlation to population density.  However, a positive correlation between cases and deaths is to be expected.  Instead, let's normalise deaths to cases and look at mortality rate instead:

```{r mortRate_graph, echo=TRUE,message=FALSE}
ggplot(data = covid_data) + geom_point(mapping = aes(x = CASES_PC, y = MORT_RATE, color = LOG_POP_DENS))
```

Even looking at mortality rate, there does not appear to be an obvious relationship to population density or case rate.

If we look at a correlation matrix instead, we can see the more objectively how well related these variables are to each other: 

```{r correlation_matrix, echo=FALSE,message=TRUE}
cor(covid_data %>% select(-STATE))
```

According to the correlation matrix, in addition to the positive relationship between deaths per capita and case rate, there also appears to be a potentially weak negative correlation between deaths per capita and population density.

## Multivariate Regression

Here we generate a multivariate linear regression model to predict deaths per capita, using only the two parameters the correlation matrix suggests have the best relationship to deaths per capita

```{r multireg_model, echo=FALSE, message=TRUE}
lm(DEATHS_PC ~ CASES_PC + POP_DENS, data = covid_data)
```

The results of the linear regression indicate that the mortality rate of COVID-19 over the course of the pandemic is approximately 8 deaths per thousand cases.  The coefficient on the population density term is small but negative, indicating that higher population densities counter-intuitively predict a lower number of deaths per capita.


## Conclusions and Biases

As seen in the correlation analysis, higher population densities are surprisingly weakly correlated to a *lower* case rate. One possible explanation for this is that perhaps densely populated areas implemented stricter or better public health measures  

There are a few potential sources of bias in this data and the statistical approach.  The most significant source of error comes from doing this analysis on the state-level, population density can vary greatly within any state, so a county level analysis might have yielded clearer results.  

An additional source of bias in the data might be political in nature.  Areas with lower-population density tend to vote more Republican, which as a party may have been reluctant to embrace government intervention during the pandemic, affecting the case numbers that are reported in their area.


